---
title: "Maximum Likelihood Estimation"
subtitle: "Social & Natural Science Applications"  
author: 
  - "Samuel Workman, Ph.D."
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [xaringan-themer.css, title_theme.css]
    nature:
      highlightStyle: solarized-light
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_solarized_light(
  header_font_google = google_font("Josefin Slab", "600"),
  text_font_google   = google_font("Alegreya Sans", "300", "300i"),
  code_font_google   = google_font("Fira Code")
)
```

class: center, middle

# The Classic Approach to Analysis

---

## Classic Procedure

1. Assume that some set of true parameters (often $\beta$) exist that describe social systems; these true parameters are fixed.

--

2. Gather some data, the observations of which are random, or in some way representative of the population.

--

3. Assume system outcomes are a function of a Gaussian distribution; note this often means contriving the data in such a way as to be continuous and to meet Gaussian assumptions.

--

4. More often than not, we parameterize and model *only* the center mass of the distribution, ignoring important variance in the dispersiion and shape. Treat the error term as statistical nuisance in the estimation of the true population parameters

--

5. Employ transformations or statistical or mathematical gimmickry to “fix” standard errors.

--

6. Discussion is restricted to the estimates of the $\beta$s which is makes explanation easy, if often tenuous.

--

7. In practice, restricts causal inference to one parameter defining one distribution within the entire realm of probability distributions.

---

## Problems with the Classical Approach

1. The data are not random...they are the result of a social or natural system and a data generating process (DGP).

--

2. The approach to the stochastic component belies the fact that the data is in part a product this stochastic component.

--

3. The normal regression model is itself based on 2 parameters; often times, the second parameter is the most interesting (e.g., variation across MCs, states, school systems, or countries).

--

4. Most interesting social phenomena are not normally distributed or continuous (e.g., votes, protests, even budgets).

--

5. Massaging data until it meets the assumptions of the classic approach means layering measurement error onto slack in model specification; also means that explanations are not intuitive and tortured (e.g., people simply do not think in terms of logged dollars).

--

6. At worst, the model estimates precisely a *true* $\beta$ that generates nonsensical substantive implications.

--

7. Fails to incorporate stochastic uncertainty into substantive implications of the model and counterfactual analysis.

---

## Key Departures in Likelihood

* Bayesian about data—the data are the data, not some random draw
from a population with *true* parameters.

--

* There are no recoverable *true* population parameters---parameters are a product of a distribution, itself a result of a social or natural data-generating process.

--

* In other words, the data, not the parameters, are fixed.

--

* Makes the stochastic component of the data-generating process integral to understanding the dependent variable. The real world is not a nuisance.

--

* Likelihood treats uncertainty as relative, rather than as an objective quantity (Bayesian about uncertainty and learning).

--

* Use social science theory and the data generating process to derive the stochastic component of our models directly.

---

class: left, middle

# Next - Deriving the Likelihood for Gaussian Outcomes

# .blue[Up-ahead - Maximum Likelihoood Methods for Least Squares]

---

class: center, middle

# Samuel Workman, Ph.D.

### .red[`r icon::fa("envelope")`] [samuel.workman@ou.edu](mailto:samuel.workman@ou.edu)

### .blue[`r icon::fa("globe-americas")`] [samuelworkman.org](https://www.samuelworkman.org)

### .green[`r icon::fa("medium")`] [@samuel.workman](https://medium.com/@samuel.workman)

### .blue[`r icon::fa("twitter")`] [@SamuelGWorkman](https://twitter.com/SamuelGWorkman)