---
title: "Maximum Likelihood Estimation"
subtitle: "Social & Natural Science Applications"  
author: 
  - "Samuel Workman, Ph.D."
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [xaringan-themer.css, title_theme.css]
    nature:
      highlightStyle: solarized-light
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_solarized_light(
  header_font_google = google_font("Josefin Slab", "600"),
  text_font_google   = google_font("Alegreya Sans", "300", "300i"),
  code_font_google   = google_font("Fira Code")
)
```

class: center, middle

# Mathematics & Probability for Likelihood

---

## Understanding Sets

* A *set*, $\mathcal{S}$, is simply a collection of elements $\{s_1, s_2,...,s_i\}$, formally:

\begin{align}
\mathcal{S} = \{s_1, s_2,...,s_i\}
\end{align}

--

* Element $s_i$ is an element of, or "in", *set* $\mathcal{S}$:

\begin{align}
s_i \in \mathcal{S}
\end{align}

--

* $\mathcal{S}$ is an empty set if it contains no elements:

\begin{align}
\mathcal{S} = \emptyset
\end{align}
---

## Subsets & Relations Among Sets

* $\mathcal{S}$ is a **subset** of the larger set $\mathcal{T}$:

\begin{align}
\mathcal{S} \subset \mathcal{T}
\end{align}

--

* The **union** of sets $\mathcal{S}$ and $\mathcal{T}$ is the new set containing all elements in at least one of the sets; in other words, all elements not outside the sets

\begin{align}
\mathcal{S} \cup \mathcal{T}
\end{align}

--

* The **intersection** of sets $\mathcal{S}$ and $\mathcal{T}$ is the new set containing all elements in both of the sets

\begin{align}
\mathcal{S} \cap \mathcal{T}
\end{align}

---

## Defining Probability Models

Probability models are defined by 3 simple assumptions for a given sample
space $\Omega$, containing all possible sets:

--

* The probability of any possible event from the social process lies between zero and one. 

--

\begin{align}
Pr(a_i) \in [0:1] \text{ } \forall \text{ } i
\end{align}

--

* Something will happen with probability one.

--

\begin{align}
Pr(\mathcal{\Omega}) = 1
\end{align}

--

* The probability of unions of $n$ pairwise disjoint events is the sum of their individual probabilities. So, if $\{a_1, a_2, \ldots, a_i\}$ are mutually exclusive, then the additive property applies, such that:

--

\begin{align}
Pr(a_1 \cup a_2 \cup a_i) = Pr(a_1) + Pr(a_2) + \ldots + Pr(a_i)
\end{align}

---

## Calculating Probabilities

* Probability of Unions

\begin{align}
Pr(A \cup B) = Pr(A) + Pr(B) - Pr(A \cap B)
\end{align}

--

* Probability of Intersections

\begin{align}
Pr(A \cap B) = Pr(A) + Pr(B) - Pr(A \cup B)
\end{align}

--

* Probability of Complements 

\begin{align}
Pr(A^C) = 1 - Pr(A) \\
Pr(A) = 1 - Pr(A^C)
\end{align}

--

* Probability of Null Set

\begin{align}
Pr(\emptyset) = 0
\end{align}

--

* Probability of Sample Space 

\begin{align}
Pr(A) = 1
\end{align}

---

## Probability as a Basis for Modeling Uncertainty

* **marginal probability** - the probability of an event irrespective of the values of another variable.

\begin{align}
Pr(X = A)
\end{align}

--

* **joint probability** - the probability of any number of events occurring, $Pr(X = A)$ *and* $Pr(X = B)$

\begin{align}
Pr(A \cap B) = Pr(AB)
\end{align}

--

* **conditional probablity** - the probability of an event given another event, $Pr(A|B)$

$$\text{conditional probablity} = \frac{\text{joint probability}}{\text{marginal probability}}$$
$$\frac{Pr(A \cap B)}{Pr(B)}$$
---

## Using Conditional Probability to Derive Bayes Rule

The probability of $a$ given $b$ depends on the probability of $b$ given $a$ and on the marginal probabilities of $a$ and $b$.

$$\text{conditional probablity} = \frac{\text{joint probability}}{\text{marginal probability}}$$

--

\begin{align}
Pr(a|b) &= \frac{Pr(a \cap b)}{Pr(b)} \\
Pr(b|a) &= \frac{Pr(b \cap a)}{Pr(a)} \\
Pr(b|a)Pr(a) &= Pr(b \cap a) \\
Pr(a|b)Pr(b) &= Pr(a \cap b) \\
Pr(a|b)Pr(b) &= Pr(b|a)Pr(a) \\
Pr(a|b) &= \frac{Pr(b|a)Pr(a)}{Pr(b)}
\end{align}

---

## Notes on Exponents

* Zero Property: $x^0 = 1$

--

* One Property: $x^1 = x$

--

* Fraction Property: $(\frac{x}{y})^a = (\frac{x^a}{y^a}) = x^ay^{-a}$

--

* Nested Exponents: $(x^a)^b = x^{ab}$

--

* Distributive Property: $xy^a = x^ay^a$

--

* Product Property: $x^ax^b = x^{a+b}$

--

* Ratio Property: $x^{\frac{a}{b}} = (x^a)^{\frac{1}{b}} = (x^{\frac{1}{b}})^a = \sqrt[b]{x^a}$

---

## Notes on Logarithms

* Zero--One Property: $log_b(1) = 0$

--

* Base & Identity: $log_b(b^x) = x$

--

* Multiplication: $log(xy) = log(x) + log(y)$

--

* Division: $log(\frac{x}{y}) = log(x) - log(y)$

--

* Exponentiation: $log(x^y) = y log(x)$

---

class: left, middle

# Next - The Classical Approach to Social Research

# .blue[Up-ahead - Maximum Likelihood Methods]

---

class: center, middle

# Samuel Workman, Ph.D.

### .red[`r icon::fa("envelope")`] [samuel.workman@ou.edu](mailto:samuel.workman@ou.edu)

### .blue[`r icon::fa("globe-americas")`] [samuelworkman.org](https://www.samuelworkman.org)

### .green[`r icon::fa("medium")`] [@samuel.workman](https://medium.com/@samuel.workman)

### .blue[`r icon::fa("twitter")`] [@SamuelGWorkman](https://twitter.com/SamuelGWorkman)