---
title: "Maximum Likelihood Estimation"
subtitle: "Social & Natural Science Applications"  
author: 
  - "Samuel Workman, Ph.D."
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [xaringan-themer.css, title_theme.css]
    nature:
      highlightStyle: solarized-light
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_solarized_light(
  header_font_google = google_font("Josefin Slab", "600"),
  text_font_google   = google_font("Alegreya Sans", "300", "300i"),
  code_font_google   = google_font("Fira Code")
)
```

class: center, middle

# The Classical Approach to Social Research

---



---

class: left, middle

# Next - Deriving the Likelihood for Gaussian Outcomes

# .blue[Up-ahead - Maximum Likelihoood Methods for Least Squares]

---

class: center, middle

# Samuel Workman, Ph.D.

### .red[`r icon::fa("envelope")`] [samuel.workman@ou.edu](mailto:samuel.workman@ou.edu)

### .blue[`r icon::fa("globe-americas")`] [samuelworkman.org](https://www.samuelworkman.org)

### .green[`r icon::fa("medium")`] [@samuel.workman](https://medium.com/@samuel.workman)

### .blue[`r icon::fa("twitter")`] [@SamuelGWorkman](https://twitter.com/SamuelGWorkman)