---
title: "Maximum Likelihood Estimation"
subtitle: "Social & Natural Science Applications"  
author: 
  - "Samuel Workman, Ph.D."
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    includes:
      in_header: "in_header.html"
      after_body: "after_body.html"
    css: [xaringan-themer.css, title_theme.css]
    nature:
      highlightStyle: solarized-light
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_solarized_light(
  header_font_google = google_font("Josefin Slab", "600"),
  text_font_google   = google_font("Alegreya Sans", "300", "300i"),
  code_font_google   = google_font("Fira Code")
)
```

class: center, middle

# Introduction & Course Content

---

## Course Topics

1. Logic of Bayesian Inference (with no informative prior, Bayesian estimates collapse to the MLE estimate)

--

2. Probability model of uncertainty; propagating this uncertainty into
substantive, model-based inference

--

3. Focus on understanding the construction of distributions and how we
might parameterize these

--

4. Be Bayesian about data...

--

5. Understand how data generating processes and the construction of measurement dictates, or bounds, modeling strategies

--

6. Using `r icon::fa("r-project")` as the core tool of analysis

---

## Persistent Themes

1. Attention to fit of the research design, and statistical procedures, to our research questions

--

2. Attention to data generating process as the central guide to modeling
strategies

--

3. Conceptual understanding of the underlying math for MLE

--

4. Model-based simulation in order to more fully understand statistical
procedures, inference, and substantive implications of model findings

--

5. Visualizing data---for descriptive purposes, to understand causal inference, the robustness of our models and measures, and the
substantive impact of model findings; communicating research results to
a general audience

--

6. Applied data analysis --- lots of `r icon::fa("r-project")` Code

---

class: center, middle

# Vignettes & Research Paper

---

## Content of the Vignettes

* Vignettes should explain a method or statistical procedure in a way that a generalist in data science, analytics, or statistics can understand

--

* Vignettes should contain `r icon::fa("r-project")` code, but also use *text* to explain the chunks and lines of code - key difference between the vignettes and the research paper

--

* Vignettes should use real data and cover all coding and statistical procedures necessary to procure or collect the data, clean the data, analyze the data, and communicate the results

--

* Vignettes do not contain literature reviews, social science theory, or broad discussions of implications; they may illustrate inference however, or how to learn from data or an analytical procedure

--

* See ["Scraping, Downloading, and Storing PDFs in R" in *Towards Data Science*](https://towardsdatascience.com/scraping-downloading-and-storing-pdfs-in-r-367a0a6d9199) for an example of a vignette.

---

## Content of the Research Paper

* In general, the paper should look like a *journal submission*

--

* The research design should map onto the research question, theory, and
hypotheses; **this is the most common reason for rejection at journals**

--

* The paper will contain the *mathematical representation* of the model
and other statistical procedures

--

* The paper will contain visual displays of quantitative information
(VDQIs)---for both descriptive and causal inference

--

* Serious effort to engage *counterfactual* scenarios and explanations

---

## Structure of the Research Paper

1. Research question and why it is important—no talk of "gap" filling.

--

2. Theoretical (causal models) & conceptual development (measurement
models).

--

3. Data & Measures---justification of indicators and measures, noting
potential measurement errors or conceptual and theoretical strengths.

--

4.  Hypotheses---map theoretical expectations and concepts directly onto
measures.

--

5. Methodological Approach---including mathematical and theoretical
justification for statistical models and procedures.

--

6. Findings---must contain discussion of robustness of approach and
estimates.

--

7. Discussion---must contain VDQIs and exploration of counterfactuals;
emphasize substantive findings.

--

8. Conclusion---what did we learn...theoretically, conceptually,
methodologically...about the world.

---

class: left, middle

# Next - Review of Mathematics for Likelihood

# .blue[Up-ahead - The Classical Approach to Social Science Modeling]

---

class: center, middle

# Samuel Workman, Ph.D.

### .red[`r icon::fa("envelope")`] [samuel.workman@ou.edu](mailto:samuel.workman@ou.edu)

### .blue[`r icon::fa("globe-americas")`] [samuelworkman.org](https://www.samuelworkman.org)

### .green[`r icon::fa("medium")`] [@samuel.workman](https://medium.com/@samuel.workman)

### .blue[`r icon::fa("twitter")`] [@SamuelGWorkman](https://twitter.com/SamuelGWorkman)